---
title: "Lab 18: Pertussis Mini Project"
author: "Rahul Nedunuri (PID: A16297840)"
format: pdf
editor: visual
---

## 1. Investigsting Pertussis Cases by Year

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.


```{r}
library(ggplot2)
library(datapasta)
```
```{r}
library(dplyr)
```


```{r}
cdc <- data.frame(year = c(1922L,1923L,1924L,1925L,
                           1926L,1927L,1928L,1929L,1930L,1931L,
                           1932L,1933L,1934L,1935L,1936L,
                           1937L,1938L,1939L,1940L,1941L,1942L,
                           1943L,1944L,1945L,1946L,1947L,
                           1948L,1949L,1950L,1951L,1952L,
                           1953L,1954L,1955L,1956L,1957L,1958L,
                           1959L,1960L,1961L,1962L,1963L,
                           1964L,1965L,1966L,1967L,1968L,1969L,
                           1970L,1971L,1972L,1973L,1974L,
                           1975L,1976L,1977L,1978L,1979L,1980L,
                           1981L,1982L,1983L,1984L,1985L,
                           1986L,1987L,1988L,1989L,1990L,
                           1991L,1992L,1993L,1994L,1995L,1996L,
                           1997L,1998L,1999L,2000L,2001L,
                           2002L,2003L,2004L,2005L,2006L,2007L,
                           2008L,2009L,2010L,2011L,2012L,
                           2013L,2014L,2015L,2016L,2017L,2018L,
                           2019L,2020L,2021L),
         cases = c(107473,164191,165418,152003,
                   202210,181411,161799,197371,
                   166914,172559,215343,179135,265269,
                   180518,147237,214652,227319,103188,
                   183866,222202,191383,191890,109873,
                   133792,109860,156517,74715,69479,
                   120718,68687,45030,37129,60886,
                   62786,31732,28295,32148,40005,
                   14809,11468,17749,17135,13005,6799,
                   7717,9718,4810,3285,4249,3036,
                   3287,1759,2402,1738,1010,2177,2063,
                   1623,1730,1248,1895,2463,2276,
                   3589,4195,2823,3450,4157,4570,
                   2719,4083,6586,4617,5137,7796,6564,
                   7405,7298,7867,7580,9771,11647,
                   25827,25616,15632,10454,13278,
                   16858,27550,18719,48277,28639,32971,
                   20762,17972,18975,15609,18617,
                   6124,2116)
       )


head(cdc)
```

```{r}
ggplot(cdc) +
  aes(year, cases) +
  #geom_point() +
  geom_line() +
  labs(x='Year', y='Number of Cases', title="Pertussis Cases by Year (1922-2021)")
```

## 2. A tale of two vaccines (wP & aP)

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) +
  aes(year, cases) +
  #geom_point() +
  geom_line() +
  labs(x='Year', y='Number of Cases', title="Pertussis Cases by Year (1922-2021)") +
  geom_vline(xintercept=1947, color='blue') + 
  geom_vline(xintercept=1992, color='red') + 
  geom_vline(xintercept=2020, color='orange')

```


> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

Immune system efficacy of younger generation wanes after around 10 years as against the older generations who received the wP vaccine.

## 3. Exploring CMI-PB data

Central question: Why does aP immunity wane faster than wP immunity?

Accessing Data from CMI-PB project

This database (like many modern projects) uses an API to return JSON format data. 
We will use the R package `jsonlite`
```{r}
library(jsonlite)
```

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```
58 wP subjects
60 aP subjects

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```
39 males
79 females

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```
## Side note: working with dates

We can use the lubridate package to make using dates easier.
`ymd` : year-month-date
`mdy` : month-day-year

```{r}
library(lubridate)
```

```{r}
today()
```
```{r}
time_length( today() - ymd("2002-01-04"),  "days")
```
I am 8098 days old.

So what is the age of everyone on our dataset?

```{r}
subject$age <- time_length(today() - ymd(subject$year_of_birth), "years")
head(subject)
```

Let's verify that wP subjects are generally older than aP subjects.

```{r}
ggplot(subject) +
  aes(age, fill=infancy_vac) +
  geom_histogram()
```

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
ap <- subject %>% filter(infancy_vac == "aP")

round( summary(ap$age) )
```
The average age of aP individuals is 26.


```{r}
wp <- subject %>% filter(infancy_vac == "wP")

round( summary(wp$age) )
```
The average age of wP individuals is 37.

```{r}
t.test(wp$age, ap$age)
```

Based on a t test, the difference in age of aP and wP ages is statistically significant at the 99% confidence level. 


> Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```



> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

Yes, with the faceted histogram, the ages of the two groups looks significantly different. 
```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```



## Get more data from CMI-PB

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = T)
head(specimen)
```

We need to **join** the specimen and subject tables to make a single meta table using `dplyr` join functions.

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:



```{r}
meta <- inner_join(subject, specimen)
head(meta)
```

Now lets read some of the other data
```{r}
ab_titer <- read_json("https://www.cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = TRUE) 
head(ab_titer)
```

Let's join the titer data to the existing meta dataframe using another call to `inner_join()`

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(meta, ab_titer)
head(abdata)
```


> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```
Isotype:        IgE  IgG IgG1 IgG2 IgG3 IgG4 
Entry counts: 6698 3233 7961 7961 7961 7961 


> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```
The most recent dataset has the fewest number of rows (the year 2022 has 2170 rows)


```{r}
table(abdata$antigen)
```

Let's make a boxplot of antigen types (y) and MFI (x).
```{r}
ggplot(abdata) +
  aes(MFI, antigen) +
  geom_boxplot()
```

Why are certain antigens very variable?

Can you facet or even color the plot by aP vs wP (infancy_vaccination)?

```{r}
ggplot(abdata) +
  aes(MFI, antigen, col=infancy_vac) +
  geom_boxplot() 
```


> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
igg <- abdata %>% filter(isotype == "IgG")
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
  xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```


 >Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

There are potentially some differences in antigen levels but in general it is hard to tell with this whole dataset overview. Mainly we see differences in the FIM2/3, PT, and FHA MFI levels over time because these are included in the vaccine. 

> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```
```{r}
filter(igg, antigen=="PRN") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(igg, antigen=="PT") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

> Q16. What do you notice about these two antigens time courses and the PT data in particular?

It appears that PT and FIM2/3 levels clearly rise over time. Whereas PRN appears to not be significantly changing over the duration of the visits. 



```{r}
table(abdata$dataset)
```


Let's focus on just the 2021_dataset

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

table(abdata.21$dataset)

```

Focus on PT antigen IgG levels. 
```{r}
pt.21 <- filter(abdata.21, isotype == "IgG", antigen == "PT")

```

Plot of planned days relative to boost and normalized MFI.
```{r}
ggplot(pt.21) +
  aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
    labs(x="Time in days after boost", y="MFI normalized levels")
```
wP individuals have a higher peak 14 days post vaccination. 


> Q17. Do you see any clear difference in aP vs. wP responses?

There is no clear difference in aP vs wP responses in terms of the PRN, PT, and FIM2/3 antigen levels with the boxplot regarding each visit, although the lineplot detailing time (in days) shows that wP individuals have a higher peak 14 days post vaccination. 


