---
title: "Class 16 Zebrafish TSV"
author: "Rahul Nedunuri (PID: A16297840)"
format: pdf
editor: visual
---

## TSV file
class16/mm-second.x.zebrafish.tsv


```{r}
library(ggplot2)
library(bio3d)


df <- read.table(file = 'mm-second.x.zebrafish.tsv', sep = '\t', header = F)
colnames(df) <- c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")
head(df)

```

Histogram
```{r}
ggplot(df, aes(x=bitscore), breaks = 30) + 
  geom_histogram(binwidth=10)
```
> What do you notice here? Note that larger bitscores are better.

I notice that the distribution looks more or less bimodal, with peaks near 30 and then around 275. There is also a skewness with a few very high bitscores above 750.

```{r}
ggplot(df, aes(pident, bitscore)) + geom_point(alpha=0.1)
```

> Is there a straightforward relationship between percent identity ($pident) and bitscore ($bitscore) for the alignments we generated?

There appears to be a positively correlated relationship between pident and bitscore. We notice, in general, that larger pidents tend to be associated with larger bitscores. 

```{r}
ggplot(df, aes((df$pident * (df$qend - df$qstart)), bitscore)) + geom_point(alpha=0.1) + geom_smooth()

```

